---
title: "HIPC ImmuneSignatures Project Meta Analysis Using Data from ImmuneSpace"
authors: "HIPC ImmuneSignatures Project Collaborators"
package developer: "Evan Henrich, Gottardo Lab @ Fred Hutchinson Cancer Research Center"
contact: "ehenrich@fredhutch.org"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

***
## PACKAGE INFORMATION
#### Created: May 2017
#### Study Authors: HIPC ImmuneSignature Collaborators
#### Package Developer: Evan Henrich (Gottardo Lab - FHCRC)
#### Contact: ehenrich@fredhutch.org

***

## GENERAL NOTES:
This markdown file outputs the results of the HIPC ImmuneSignatures Project using the original parameters and
streamlined code from the published manuscript but run with data coming directly from ImmuneSpace. In order 
to save time, the gene expression matrices are pre-loaded as data within the package.  The code to generate 
these expression matrices from ImmuneSpace / GEO can be found in the "get_GE.R" script within the package.
There are two cases where data from ImmuneSpace could not be used: SDY80 (hai) and SDY400 (ge).  

The SDY80 hai data available on ImmuneSpace is not incorporated into the analysis because it does not include 
subjects without gene expression data.  Although subjects without GE data are filtered out later, they must 
be included in the adjust_hai.R script in order to achieve the same results for the "endPoint" variable 
("fc_res_maxd30").  The original raw HAI data came from Yuri Kotliarov at NIH and is included within the package. 

The SDY400 gene expression data used in the package comes from the authors at Yale University because the "series.txt"
file on GEO has different probe-to-gene symbol mapping that cannot be retroactively changed, while the
"non-normalized.txt" version on GEO has experiment specific subject ID mappings that were not able to be correctly
given ImmuneSpace subject ids at the time of this report's development.

Finally, the results for the "old" cohort using SDY67 data are slightly different ( <0.5% Relative Error ).  This is 
is due to rounding that occurs from a transformation from one kind of data structure to another.

## PACKAGE NOTES:
In order to run this vignette locally, you would first need to install the following Bioconductor packages by hand: Biobase, qusage, and MetaIntegrator. Then install `ImmSig2` via `devtools::install_github("ehfhcrc/ImmSig2", build_vignettes=T)` and view the vignette with `vignette("ImmSig_Pipeline_demo", package="ImmSig2")`.  

All original code developed for the production of the manuscript is also included in the /OrigCode folder within the package.  This can be viewed at <https://github.com/ehfhcrc/ImmSig2>. 

```{r set-options, include=FALSE}
# Eset prep
library(ImmSig2) # to load data and helper functions
library(Rlabkey)
library(Biobase) # biocLite
library(stringr)
library(hash)
library(dplyr)
# library(RCurl) # used by get_GE and data preloaded so not using
library(data.table)
# library(preprocessCore) # biocLite, used in get_GE

# Module / Gene Set Analysis
library(qusage) # biocLite

# Single Gene Analysis
library(pROC) 
# library(RMySQL) # depends on libmysqlclient / used in Single Gene Analysis once, data preloaded here
# sudo apt-get install libmysqlclient-dev
library(ggplot2) 
library(clinfun) 
library(MetaIntegrator) # biocLite

options(width = 999)
knitr::opts_chunk$set(echo=FALSE, 
               # cache.path=paste(labkey.file.root, "cache/pubmed_report_cache/", sep="/"), # for IS Report
               cache=TRUE, 
               warning=FALSE, 
               message=FALSE, 
               tidy=TRUE,
               fig.height=7, 
               fig.width=7,
               fig.align = "center")
```

#### Generating and Pre-Processing expressionSet Objects

Construct expressionSet objects from ImmuneSpace data and put into a list for use in the analysis pipeline.

```{r prepare-esets, echo=TRUE}
studies <- c("SDY212","SDY63","SDY404","SDY400","SDY67","SDY80")
eset_list <- list()

for(sdy in studies){
  if(sdy == "SDY400"){
    # SDY400:
    # Using original table because GEO series matrix produces different results in terms of gene symbols
    # incorporated into qusage analysis, perhaps due to different normalization method(?), and GEO
    # non-normalized.txt does not have sample names that can be mapped with available information.
    # All original tables come from HIPC Google Drive.
    em <- get(paste0(tolower(sdy), "_exprMxOrig"))
  } else {
    # See comments in get_GE.R for methodology of generating expression matrices. These matrices
    # are created using online, public sources (e.g. GEO or ImmuneSpace).
    em <- getGE(sdy) 
  }
  
  # Remove dup biosamples based on Yale HIPCMetaModuleAnalysis.R comments
  if(sdy == "SDY212"){ em <- em[ , -(grep("SUB134307.*", colnames(em))) ] }
  
  # Yale collaborators have not yet updated GEO database to reflect these changes
  if(sdy == "SDY404"){
    em <- swapCols(em, "SUB120473_d0", "SUB120474_d0") # Incorrect Sex
    em <- swapCols(em, "SUB120460_d0", "SUB120485_d28") # Strong evidence
  }

  new_fData <- data.frame(gene_symbol = em$geneSymbol, stringsAsFactors = F)
  rownames(new_fData) <- rownames(em)
  eset <- ExpressionSet(assayData = as.matrix(em),
                        featureData = AnnotatedDataFrame(new_fData))

  # Parse Subject IDs and Limit expression data to baseline (d0) only
  eset <- eset[ , grep("d0", sampleNames(eset)) ]

  # Get baseline eset Subjects
  factored_ids <- str_match(sampleNames(eset), "(((SUB)|(sub))_?[0-9]+)(\\.[0-9])?_(.*)")
  eset_subs <- factored_ids[, 2L]

  # Get raw HAI data from Immunespace where feasible and adjust according to Yuri Kotliarov code.
  # NOTE: SDY80 raw data in ImmuneSpace differs, therefore using preloaded.
  con <- CreateConnection(sdy)
  raw_hai <- if(sdy != "SDY80"){ con$getDataset("hai") }else{ SDY80_rawtiterdata_v2 }
  adj_hai <- adjust_hai(sdy, raw_hai)

  # Age needed for SDY80 to remove subjects over 36 years old
  if(sdy == "SDY80"){
    demo <- con$getDataset("demographics")
    adj_hai$Age <- sapply(adj_hai$subject, FUN = function(x){
      trg <- demo[ which(demo$participant_id == paste0(x, ".80"))]
      return(trg$age_reported)
    })
    adj_hai <- adj_hai[ which(adj_hai$Age < 36), ]
  }

  # Ensure HAI and eset contain same subjects in same order, then push to pData
  eset <- eset[ , eset_subs %in% adj_hai$subject ]
  adj_hai <- adj_hai[ adj_hai$subject %in% eset_subs, ]
  adj_hai <- adj_hai[ order(match(adj_hai$subject, eset_subs)), ]
  rownames(adj_hai) <- sampleNames(eset)
  pData(eset) <- droplevels(adj_hai)
  
  # push to holder
  eset_list[[sdy]] <- eset
}

# Prep GeneSetDB per original code.  Can also use GSEAbase package.
gene_set <- strsplit(geneSetDB, "\t")            ## convert from vector of strings to a list
names(gene_set) <- sapply(gene_set,"[", 1)       ## move the names column as the names of the list
links <- sapply(gene_set,"[", 2)
gene_set <- lapply(gene_set, "[",-1:-2)         ## remove name and description columns
gene_set <- lapply(gene_set, function(x){ x[ which( x != "") ] })      ## remove empty strings
```

The list of expressionSet objects are then adjusted for each cohort to select the correct gene expression and hai information by subsetting the subjects.  SDY80 is treated differently in terms of the hai data and noted below.

```{r adjust-esets-by-cohort, echo=TRUE}
adjust_eset <- function(sdy, eset_list, cohort){
  eset <- eset_list[[sdy]]
  storageMode(assayData(eset)) <- "list" ## change from "lockedEnvironment" to modify assayData
  
  ## Select correct PhenoData in terms of young or old fc_res_max_d30 and delete rest
  hai_var <- grep("^((fc)|(d0))_", varLabels(eset), value = TRUE) 
  oppo <- ifelse( cohort == 'young', "old", "young")
  eset <- eset[, eset$Age.class %in% cohort] # set in adjust_hai
  
  # Per notes from F. Vallania @ Stanford and H. Meng @ Yale, SDY80 uses the adjMFC or fc_res_max_d30 
  # calculated from a combined group of young and slightly older subjects because it improved the estimation
  if(sdy != "SDY80"){
    phenoData(eset) <- phenoData(eset)[, setdiff(varLabels(eset),
                                               c(hai_var, paste0(oppo, "_", hai_var)))]
    varLabels(eset) <- gsub(paste0("^", cohort, "_"), '', varLabels(eset))
  }else{
    hai_var_y <- paste0("young_", hai_var)
    hai_var_o <- paste0("old_", hai_var)
    phenoData(eset) <- phenoData(eset)[ , !(colnames(phenoData(eset)) %in% c(hai_var_o, hai_var_y))]
  }
  
  # remove NAs from fc_res_max_d30 pheno and therefore exprs to streamline
  eset <- eset[ , !is.na(eset$fc_res_max_d30) ] 
  
  storageMode(assayData(eset)) <- "lockedEnvironment"
  return(eset)
}

esetLsMkr <- function(eset_list, cohort){
  valSdy <- ifelse(cohort == "young", "SDY80", "SDY67")
  corrEsetList <- eset_list[ c("SDY212","SDY63","SDY404","SDY400", valSdy) ]
  adjEsetList <- lapply(names(corrEsetList), adjust_eset, 
                        eset_list = corrEsetList, 
                        cohort = cohort)
  names(adjEsetList) <- names(corrEsetList)
  return(adjEsetList)
}

yngEsetLs <- esetLsMkr(eset_list, cohort = "young")
oldEsetLs <- esetLsMkr(eset_list, cohort = "old")
```

## GENE SET META-ANALYSIS
***
The Gene Set or Module meta analysis based on work by the Yale HIPC collaborators is run first.  The analysis is performed by the `QuSage` package that is found on BioConductor.

### YOUNG COHORT RESULTS
```{r meta-analysis-young, include=TRUE, results="markup"}
yng_res <- meta_analysis(adj_eset_list = yngEsetLs,
                         cohort = "young",
                         gene_set = gene_set)
```

#### Discovery Group Significant Pathways Data  
``` {r include=TRUE}
DT::datatable(yng_res$dsc)
```

#### Validation Study Significant Pathways Data  
``` {r include=TRUE}
DT::datatable(yng_res$val)
```

***

### OLD COHORT RESULTS

```{r meta-analysis-old, include=TRUE, results="markup"}
old_res <- meta_analysis(adj_eset_list = oldEsetLs,
                         cohort = "old",
                         gene_set = gene_set)
```

#### Discovery Group Significant Pathways Data  
``` {r}
DT::datatable(old_res$dsc)
```

#### Validation Study Significant Pathways Data  
``` {r}
DT::datatable(old_res$val)
```

***

#### PARAMETERS:
- **FDR.cutoff: ** 0.5
- **pvalue.cutoff: ** 0.0105
- **endPoint: ** fc_res_max_d30
- **geneSetDB: ** BTM_for_GSEA_20131008.gmt

***

## SINGLE GENE ANALYSIS
The single gene meta analysis was created by the Stanford HIPC collaborators and uses their package `MetaIntegrator`, which is also available on BioConductor.  

```{r Single-Gene-Analysis, include=FALSE}
  sgaResYng <- singleGeneAnalysis(adjEsetList = yngEsetLs, cohort = "young")
```

### YOUNG COHORT
***
#### Discovery - Upregulated Genes
```{r}
  DT::datatable(sgaResYng$disc$upReg)
```

#### Discovery - Downregulated Genes
```{r}
  DT::datatable(sgaResYng$disc$dnReg)
```

***
#### Violin Plot - SDY80
```{r}
  sgaResYng$val$ViolinPlot
```

#### ROC Plot - SDY80
```{r}
  sgaResYng$val$RocPlot
```
***

### OLD COHORT
No Significant up or down-regulated genes were found using the adjusted expressionSet and therefore analysis is not performed on SDY67.
