---
title: "HIPC ImmuneSignatures Project Meta Analysis Using Data from ImmuneSpace"
authors: "HIPC ImmuneSignatures Project Collaborators"
package developer: "Evan Henrich, Gottardo Lab @ Fred Hutchinson Cancer Research Center"
contact: "ehenrich@fredhutch.org"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

***
## PACKAGE INFORMATION
#### Created: January 2017
#### Study Authors: HIPC ImmuneSignature Collaborators
#### Package Developer: Evan Henrich (Gottardo Lab - FHCRC)
#### Contact: ehenrich@fredhutch.org

***

## GENERAL NOTES:
This markdown file outputs the results of the HIPC ImmuneSignatures Project using the original parameters and
streamlined code from the published manuscript but run with data coming direclty from ImmuneSpace. Interested 
users may re-run the full pipeline with the same or other parameters using the 'hipc_full_pipeline()' method 
in the "ImmSigPkg" available on github via install_github("ehfhcrc/ImmSigPkg").  The ImmSigPkg also outputs
intermediate files similar to the original study, allowing for easier comparison through the pipeline.

```{r set-options, include=FALSE}
library(ImmSig2) # to load data and helper functions
library(Rlabkey)
library(Biobase) 
library(stringr)
library(hash)
library(dplyr)
library(qusage) 
library(org.Hs.eg.db)
library(AnnotationDbi)

options(width = 999)
knitr::opts_chunk$set(echo=FALSE, 
               cache=FALSE, 
               warning=FALSE, 
               message=FALSE, 
               tidy=TRUE,
               fig.height=7, 
               fig.width=7,
               fig.align = "center")
```

```{r include=FALSE}
# 1. Create expressionSet object using data from ImmuneSpace
studies <- c("SDY212","SDY63","SDY404","SDY400","SDY67","SDY80")
eset_list <- list()

for(sdy in studies){
  labkey.url.path <- paste0("/Studies/", sdy, "/")
  con <- CreateConnection(sdy)
  
  # Get expressionSet object with normalized gene expression values, use highest expression
  # probe for each gene_symbol and remove others. Finally map SampleNames.
  # Mapping done with "biosmpl2sub" object generated from ImmuneSpace Schema Browser.
  # NOTE 1: SDY67 was normalized using DEseq method. See ImmSigPkg code for details.
  if(sdy != "SDY67"){ # SDY67 and SDY400 not able to run through HIPCmatrix
    mats <- con$GeneExpressionMatrices()
    eset <- con$getGEMatrix(mats$name, summary = F) 
    tmp_pData <- pData(eset)
    adj_em <- adjust_GE(eset, sdy)
    new_fData <- data.frame(rownames(adj_em))
    rownames(new_fData) <- rownames(adj_em)
    colnames(new_fData) <- c("gene_symbol")
    eset <- ExpressionSet(assayData = as.matrix(adj_em),
                          featureData = AnnotatedDataFrame(new_fData),
                          phenoData = AnnotatedDataFrame(tmp_pData))
    if(sdy == "SDY212"){ eset <- eset[ , -which(sampleNames(eset) == "BS694717") ] } # Remove dup biosamples
    sampleNames(eset) <- unname(hashmap(biosmpl2sub$Biosample, biosmpl2sub$GE_ID, sampleNames(eset)))
  }else{
    eset <- get(paste0(sdy, "_min_eset"))
  } 

  # # Parse Subject IDs and Limit expression data to baseline (d0) only
  factored_ids <- str_match(sampleNames(eset), "(((SUB)|(sub))_?[0-9]+)(\\.[0-9])?_(.*)")
  eset$Condition <- factor(factored_ids[, 7L])
  eset <- eset[ , grep("^d0", eset$Condition) ]

  # Get baseline eset Subjects
  factored_ids <- str_match(sampleNames(eset), "(((SUB)|(sub))_?[0-9]+)(\\.[0-9])?_(.*)")
  eset_subs <- factored_ids[, 2L]

  # Get raw HAI data from Immunespace where feasible and adjust according to Yuri Kotliarov code.
  # NOTE: SDY80 raw data in ImmuneSpace differs, therefore using preloaded.
  if(sdy != "SDY80"){
    raw_hai <- con$getDataset("hai")
  }else{
    raw_hai <- SDY80_rawtiterdata_v2
  }

  adj_hai <- adjust_hai(sdy, raw_hai)

  # Age needed for SDY80 to remove subjects over 36 years old
  if(sdy == "SDY80"){
    demo <- con$getDataset("demographics")
    adj_hai$Age <- sapply(adj_hai$subject, FUN = function(x){
      trg <- demo[ which(demo$participant_id == paste0(x, ".80"))]
      return(trg$age_reported)
    })
    adj_hai <- adj_hai[ which(adj_hai$Age < 36), ]
  }

  # Ensure HAI and eset contain same subjects in same order, then push to pData
  eset <- eset[ , which(eset_subs %in% adj_hai$subject) ]
  adj_hai <- adj_hai[ which(adj_hai$subject %in% eset_subs), ]
  adj_hai <- adj_hai[ order(match(adj_hai$subject, eset_subs)), ]
  rownames(adj_hai) <- sampleNames(eset)
  pData(eset) <- droplevels(adj_hai)
  
  eset_list[[sdy]] <- eset
}
```

## META-ANALYSIS 

### YOUNG COHORT RESULTS
```{r include=TRUE, results="markup"}
yng_res <- meta_analysis(eset_list = eset_list,
                         cohort = "young")
```

#### Discovery Group Significant Pathways Data  
``` {r include=TRUE}
DT::datatable(yng_res$dsc)
```

#### Validation Study Significant Pathways Data  
``` {r include=TRUE}
DT::datatable(yng_res$val)
```

***

### OLD COHORT RESULTS

```{r include=TRUE, results="markup"}
old_res <- meta_analysis(eset_list = eset_list,
                         cohort = "old")
```

#### Discovery Group Significant Pathways Data  
``` {r}
DT::datatable(old_res$dsc)
```

#### Validation Study Significant Pathways Data  
``` {r}
DT::datatable(old_res$val)
```

***

### PARAMETERS:
- **FDR.cutoff: ** 0.5
- **pvalue.cutoff: ** 0.01
- **endPoint: ** fc_res_max_d30
- **geneSetDB: ** BTM_for_GSEA_20131008.gmt
