---
title: "HIPC ImmuneSignatures Project - Full analysis pipeline using data from ImmuneSpace"
authors: "HIPC ImmuneSignatures Project Collaborators"
maintainer: "Evan Henrich, Gottardo Lab @ Fred Hutchinson Cancer Research Center"
contact: "ehenrich@fredhutch.org"
created: "January 2017"
output: html_document
---
# NOTE: This markdown file is streamlined to only run the default parameter values that were
# used in the original work by the HIPC ImmuneSignatures collaborators.  However, interested 
# users may run the main function 'hipc_run_full_pipeline()' in the console and follow prompts 
# to change these parameters and see different outcomes.

```{r include=FALSE, warning=FALSE}
library(ImmSigPkg)

studies <- c("SDY212", "SDY63", "SDY404", "SDY400", "SDY80", "SDY67")

# Setup initial directory structure
ImmSig_dir <- file.path(getwd(),"ImmSig_Analysis")
dir.create(ImmSig_dir)

# 1. Extract and pre-process data from ImmuneSpace, ImmPort, or GEO databases
pp_dir <- file.path(ImmSig_dir,"PreProc_Data")
dir.create(path = pp_dir)

hai_dir <- file.path(pp_dir,"HAI")
dir.create(path = hai_dir)

ge_dir <- file.path(pp_dir,"GE")
dir.create(path = ge_dir)

rawdata_dir <- file.path(pp_dir,"rawdata")
dir.create(path = rawdata_dir)

yale.anno <- "original"
sdy80.process <- FALSE
sdy80.anno <- "original"
sdy80.norm <- FALSE
    
for(sdy in studies){
    if( !(sdy %in% c("SDY400","SDY80")) | (sdy == "SDY80" & sdy80.process == TRUE) ){
        makeGE(sdy, user, pwd,
               yale.anno = yale.anno,
               sdy80.anno = sdy80.anno,
               sdy80.norm = sdy80.norm,
               output_dir = ge_dir,
               rawdata_dir = rawdata_dir)
        makeHAI(sdy, output_dir = hai_dir)
        makeDemo(sdy, output_dir = ge_dir)
    }else if(sdy == "SDY400"){
      makeHAI(sdy, output_dir = hai_dir)
      makeDemo(sdy, output_dir = ge_dir)
    }
  }
```
#--------PreProc-Markdown-Output--------------------------------------------
********************************************************************************************
Pre-processing of HIPC Immunesignatures parameters:
## `r paste0("Yale Studies Gene Expression Annotation: ", yale.anno)
## `r paste0("Processing of SDY80/CHI-nih from IS rawdata: ", sdy80.process)
## `r paste0("SDY80/CHI-nih Gene Expression Annotation: ", sdy80.anno)
## `r paste0("SDY80/CHI-nih GE normalization via pipeleine: ", sdy80.norm)

Notes:
1. Gene expression annotation methods:
- original = using the original pre-processed GE table from HIPC collaborators to generate
  a hash of Probe IDs as keys and Gene Symbols as values.  For the Yale Studies, this method
  means that the annotation is based on the GE table provided from the Illumina sequencing
  machine at the time of the run.  For SDY80/CHI-nih, the information is from the Affymetrix
  annotation done at the time.
- library = the use of a bioconductor library.  For Yale Studies, this was illuminahumanv4.db.
  For SDY80, it was hugene10sttranscriptcluster.db.
- manifest = Uses the manifest available from the Illumina website which consists of all historic
  probeIDs.  In the case of the Yale Studies, this generates complete mapping, whereas the 
  original table has only partial mapping and the library even less.

2. SDY80 / CHI-nih normalization
- In the original code used to develop the manuscript, The study's rawdata from the GEO database 
  has been log2 transformed, but is not normalized in the same was as the discovery studies 
  (i.e. preprocessCore::normalize.quantiles).  Therefore sdy80.norm is FALSE by default.  However,
  the user may set it to TRUE and use the same normalization procedure as the other studies.

*********************************************************************************************
#---------------------------------------------------------------------------

```{r include=FALSE, warning=FALSE}

# Step 2: Combine pre-processed files into Rds (BioConductor eset)
rds_dir <- file.path(ImmSig_dir, "Rds_data")
dir.create(path = rds_dir)

combined_hai <- combine_hai_data(hai_dir, sdy80.process, output_dir = rds_dir)
for(sdy in studies){
  make_rds(sdy, ge_dir, sdy80.process, combined_hai, output_dir = rds_dir)
}

# Step 3: Run meta analysis script
output_dir <- file.path(ImmSig_dir, "meta_analysis_output")
dir.create(path = output_dir)

FDR.cutoff <- 0.5
pvalue.cutoff <- 0.01
endPoint <- "fc_res_max_d30"
adjusted <- FALSE
baselineOnly <- TRUE
indiv_rds <- FALSE
```

```{r, warning=FALSE}
meta_analysis(geneSetDB = geneSetDB,
                  rds_data_dir = rds_dir,
                  cohort = "young",
                  FDR.cutoff = FDR.cutoff,
                  pvalue.cutoff = pvalue.cutoff,
                  endPoint = endPoint,
                  adjusted = adjusted,
                  baselineOnly = baselineOnly,
                  indiv_rds = indiv_rds,
                  markdown = T,
                  output_dir = output_dir)

meta_analysis(geneSetDB = geneSetDB,
                  rds_data_dir = rds_dir,
                  cohort = "old",
                  FDR.cutoff = FDR.cutoff,
                  pvalue.cutoff = pvalue.cutoff,
                  endPoint = endPoint,
                  adjusted = adjusted,
                  baselineOnly = baselineOnly,
                  indiv_rds = indiv_rds,
                  markdown = T,
                  output_dir = output_dir)
```
#---------RMarkdown-Meta-Analysis-Output--------------------------------------------------
***************************************************************************************
Meta-Analysis of HIPC ImmuneSignatures Parameters (both cohorts - young / old:
  
## `r paste0("FDR.cutoff = ", FDR.cutoff)
## `r paste0("pvalue.cutoff = ", pvalue.cutoff)
## `r paste0("endPoint = ", Endpoint
## `r paste0("adjusted = ", adjusted
## `r paste0("baselineOnly = ", baselineOnly)
## `r paste0("indiv_rds = ", indiv_rds)
## `r paste0("geneSetDB = ", geneSetDB)

Explanation of Parameters:
- FDR.cutoff = Cut off point for q-values, in code 'combined.q <- p.adjust(combined.p, method="BH")'
- pvalue.cutoff = Cut off point for p-values when selecting gene pathways as 'significant' from the
  discovery studies. In code, 'combined.p <- pdf.pval(combinePDFs(quSageObjList, n.points = 2^14))' 
- endPoint = the HAI column used for labeling "responders" and "non-responders" to a vaccine.  The user
  has the option of fc_res_max_d30 or fc_res_max_d20, which select either the 30% / 70% points for 
  discretization or 20% / 80% respectively.
- adjusted = If TRUE, selects the age-adjusted gene expression values for use in calculations.
- baselineOnly = If TRUE, uses only day zero gene expression values in calculations.
- indiv_rds = If TRUE, outputs the rds files for the individual studies as part of the output
- geneSetDB = The file used for gene set definition.
***************************************************************************************
#-----------------------------------------------------------------------------------------

